Solar wind prediction for the Solar Probe Plus orbit

### 2017-04-08 paperMVVB/sections/content.tex
worked in 2nd part of my offline comments

\subsection{Exponential fitting}
%why exponential function?
An exponential distance behavior is expected from all four parameters (cites?). Therefore we use the exponential function
\begin{align}
	x(r) = a\,r^b	\label{eq:exponential_function}
\end{align}
with the solar distance $r$ for the regression fit of the median and mean. The fits are weighted by data counts per bin.
%fit result table and figure
With $r$ in astronomical units we get the fit coefficients ($a_\text{med}$, $a_\text{avg}$, $b_\text{med}$ and $b_\text{avg}$) as given in Table~\ref{tab:mean_median_fit_parameter}.\\
\begin{table*}
	\caption{These are the fit coefficients for the median and mean solar distance dependencies of the four parameters from the combined Helios data set. The errors in brackets are the estimated standard deviations of each fit parameter. The crossing distance is the point where median and mean intersect.}
	\label{tab:mean_median_fit_parameter}
	\centering
	\begin{tabular}{l
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	@{}c
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	S[table-format = 1.1(2)e1]}
		\hline\hline
		\multirow{2}{*}{Parameter}	&\multicolumn{2}{c}{Median}	&	&\multicolumn{2}{c}{Mean}	&\multicolumn{1}{c}{Crossing distance}\\
		\cline{2-3}	\cline{5-6}
			&\multicolumn{1}{c}{$a_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{$b_\text{med}$}	&	&\multicolumn{1}{c}{$a_\text{avg}$\tablefootmark{a}}	&\multicolumn{1}{c}{$b_\text{avg}$}	&\multicolumn{1}{c}{[\si{\au}]}\\
		\hline
		Magnetic field	&5.377(92)	&-1.655(17)	&	&6.05(10)	&-1.546(18)	&0.339(11)\\
		Velocity	&4.107(28)	&0.058(13)	&	&4.356(24)	&0.049(10)	&0.7(83)e3\\
		Density		&5.61(27)	&-2.093(46)	&	&7.57(30)	&-2.010(38)	&0.027(73)\\
		Temperature	&7.14(23)	&-0.913(39)	&	&9.67(21)	&-0.792(28)	&0.082(85)\\
		\hline
	\end{tabular}
	\tablefoot{
		\tablefoottext{a}{Values in their respective units \si{nT}, \SI{e2}{\km\per\s}, \si{\per\cm\cubed} and \SI{e4}{\K}.}
	}
\end{table*}

@@

study time/cycle variations\\

@@

Finally, we combine the obtained solar activity and distance dependencies for shifting the frequency distributions. The result will be an empirical solar wind model for the inner heliosphere.\\

%are the exponents time-independent?
Under the assumption that the exponential falloff laws do not change with time/solar activity, they can be used in general (indications for that: fig. + theory + studies/papers...).\\
assuming the distance scaling laws are time independent, which we deduced...? which is an approximation.\\

We combine the fit coefficients of the median relation for solar activity dependence (\ref{eq:median_with_ssn}) with the ones from the exponential distance dependency (\ref{eq:exponential_function}):
\begin{align}
	x_\text{med}(ssn,r) &= (a_\text{med} \cdot ssn + b_\text{med}) \cdot r^b\,.
\end{align}

the combined model function $W''(x,ssn,r)$\\

...and double lognormal function for the velocity\\

%model errors
give model errors sizes/uncertainties (sigmas); $W''_\text{err}(x,ssn,r)$

@@

SPP orbit description, Venus flybys, ecliptic, \citep{Fox2015}, see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_b_plot}\\
choosing SSN prediction source as input (SIDC + SWPC + previous cycle), (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_b_plot})\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_orbit_predicted_SSN_overview_b_plot.pdf}}
	\caption{SPP's solar distance during its mission time (top). Consecutive Venus flybys bring its perihelia nearer to the Sun. Actual and predicted SSN (bottom), i.e. 13-month smoothed monthly actual SSN and SIDC prediction, SWPC prediction and simply shifted SSN from previous cycle~24.}
	\label{fig:SPP_orbit_predicted_SSN_overview_b_plot}
\end{figure}

plot of SPP's estimated solar wind environment from predicted SSN; zoom into first and nearest perihelion (Fig.~\ref{fig:SPP_perihelia_prediction_plot})\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_plot.pdf}}
	\caption{Estimated B-field median during the first and nearest perihelia in 2018 and 2024.}
	\label{fig:SPP_perihelia_prediction_plot}
\end{figure}

@@

velocity:\\
%critical surfaces
in direction to the Sun is at $\sim$2.5~Rs the source surface (Schatten1969)\\

sonic and Alfvénic critical point positions (see \citet{Sittler1999})\\
Alfvénic critical surface i.e. source surface (see Fox before 2.1)\\

approaching these regions, acceleration plays a role\\

@@

@@

%model boundaries
model boundaries (spherical coordinates):\\
- mean solar wind of the Helios time range -> error size (determine deviation from average)\\
- heliocentric distance range 0.29--0.98~au\\
- rotational symmetry\\
- confined to ecliptic\\
model constrictions:\\
- solar distance dependency function\\
- frequency distribution function\\
- neglected influence from heliolatitude variation\\

@@

all estimates outside these boundaries are extrapolations with large uncertainties. see Sect.~\ref{sec:model_validity_and_errors}\\

@@

Generally, the use of hourly data instead of higher resolution data (less than 1~min) should not have significant influence on the results. The minutely OMNI~2 data's frequency distributions are almost congruent and differ only slightly at their extreme ends (we tested that...).\\

The many hourly Helios data points which contain only a few measurements, contribute with a larger scatter to the frequency distributions \citep[p.~491]{Schwenn1983}.\\

The OMNI~2 data is time-shifted to the nose of the Earth's bow shock. This leads to yearly solar distance variations of \SI{>2}{\percent} (cite?) as the Earth orbits the Sun. Furthermore, its orbit within the ecliptic leads to a yearly variation of \SI{+-7.2}{\degree} in heliospheric latitude.\\
The solar wind parameters vary with solar distance as well as with latitudinal separation from the heliospheric current sheet (HCS) (cite?).\\
The HCS's position in latitude is highly variable around the solar equator \citep[p.~127~ff.?]{Schwenn1990}.\\

- over year (seasonal/monthly) -> expect them to be less than 5~\%\\

%monthly seasonal variation
Sun-Earth distance over the course of the year.\\
In the year 2017 Earth's perihelion was on 5~January with a distance of \SI{-1.67}{\percent} from \SI{1}{\au} (Horizons On-Line Ephemeris System\footnote{\url{http://ssd.jpl.nasa.gov/horizons.cgi}}, Solar System Dynamics Group, Jet~Propulsion Laboratory).\\
The cosine approximation
\begin{align}
	r_\text{E}(t) = 1 - 0.0167 \cdot \cos\left(2 \pi \left(t - 2017 - \frac{5}{365}\right)\right)\,,
\end{align}
with $t$ in years, suffices for our accuracy requirements.\\
% HORIZONS Web-Interface
% http://ssd.jpl.nasa.gov/horizons.cgi
% Ephemeris Type:	VECTORS
% Target Body:		Earth [Geocenter] [399]
% Coordinate Origin:	Sun (body center) [500@10]
% Time Span:		Start=2017-01-01, Stop=2017-12-31, Step=1 d
% Table Settings:	CSV format=YES
% Display/Output:	download/save (plain text file)
%-> distance:	9.833098226363186E-01
%-> diff. to 1 au:	0.166901774
%-> change to day before:	4.4e-6	-> 0.1669018(44)

seasonal variation function:\\
$X_\text{avg}(t) = a\,r_\text{E}(t)^b$\\
% B_avg = 5.969(29)~nT\\
% B_med = 5.435(24)~nT\\
% N_avg = 6.329(85)~cm-3\\
% N_med = 4.858(60)~cm-3\\
% T_avg = 1.198(16)e5~K\\
% T_med = 7.17(12)e4~K\\
% V1_avg = 4.132(22)e2~km/s\\
% V1_med = 4.034(21)e2~km/s\\
% V2_avg = 4.132(22)e2~km/s\\
% V2_med = 4.034(21)e2~km/s\\

% Heliographic Inertial (HGI) coordinates\\
% HGI coordinates are Sun-centered with the z-axis directed along the solar rotation axis and directed northward of the solar equator. The solar equator plane is inclined \SI{7.25}{\degree} from the ecliptic.\\

HGI coordinates; latitude range -7.25\degr{}~to~7.25\degr{}\\
latitude variation (see Schwenn1990 p.~127)\\

@@

list of results:\\
- low velocity at 0.0459~au\\
- empirical solar wind model for inner heliosphere\\
- application validity of lognormal distributions\\
--> B inversion of frequency distribution\\
---> magnetic field distribution's with distance increasing high value tail -> source are compression regions (why with density no increase?); look into Parker1958's B-field formula...\\

%discussion
Comparison with existing near-Sun models (renew extrapolation figure to 1-column)\\

@@

varying shape with distance is indicator for internal physical processes (mixing/turbulence...)\\

@@

Further investigations should be done into structure extrapolations...\\

>>>>

\subsection{Exponential fitting}
%why exponential function?
An exponential distance behavior is expected from all four parameters (cites?). Therefore we use the exponential function
\begin{align}
	x(r) = a\,r^b	\label{eq:exponential_function}
\end{align}
with the solar distance $r$ for the regression fit of the median and mean. The fits are weighted by data counts per bin.
%fit result table and figure
With $r$ in astronomical units we get the fit coefficients ($a_\text{med}$, $a_\text{avg}$, $b_\text{med}$ and $b_\text{avg}$) as given in Table~\ref{tab:mean_median_fit_parameter}.\\
\begin{table*}
	\caption{These are the fit coefficients for the median and mean solar distance dependencies of the four parameters from the combined Helios data set. The errors in brackets are the estimated standard deviations of each fit parameter. The crossing distance is the point where the fitted median and mean intersect.}
	\label{tab:mean_median_fit_parameter}
	\centering
	\begin{tabular}{l
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	@{}c
	S[table-format = 1.3(2)]
	S[table-format = +1.3(2)]
	S[table-format = 1.1(2)e1]}
		\hline\hline
		\multirow{2}{*}{Parameter}	&\multicolumn{2}{c}{Median}	&	&\multicolumn{2}{c}{Mean}	&\multicolumn{1}{c}{Crossing distance}\\
		\cline{2-3}	\cline{5-6}
			&\multicolumn{1}{c}{$a_\text{med}$\tablefootmark{a}}	&\multicolumn{1}{c}{$b_\text{med}$}	&	&\multicolumn{1}{c}{$a_\text{avg}$\tablefootmark{a}}	&\multicolumn{1}{c}{$b_\text{avg}$}	&\multicolumn{1}{c}{[\si{\au}]}\\
		\hline
		Magnetic field	&5.377(92)	&-1.655(17)	&	&6.05(10)	&-1.546(18)	&0.339(11)\\
		Velocity	&4.107(28)	&0.058(13)	&	&4.356(24)	&0.049(10)	&0.7(83)e3\\
		Density		&5.61(27)	&-2.093(46)	&	&7.57(30)	&-2.010(38)	&0.027(73)\\
		Temperature	&7.14(23)	&-0.913(39)	&	&9.67(21)	&-0.792(28)	&0.082(85)\\
		\hline
	\end{tabular}
	\tablefoot{
		\tablefoottext{a}{Values in their respective units \si{nT}, \SI{e2}{\km\per\s}, \si{\per\cm\cubed} and \SI{e4}{\K}.}
	}
\end{table*}

@@

study time/cycle variations\\
indications for that: fig. + theory + studies/papers...\\
assuming the distance scaling laws are time independent, which we deduced...? which is an approximation.\\

@@

Finally, we combine the obtained solar activity and distance dependencies for shifting the frequency distributions. The result is an empirical solar wind model for the inner heliosphere.\\

%are the exponents time-independent?
Under the assumption that the exponential falloff laws do not change with time/solar activity (as shown above...), they can be used in general.\\

We combine the fit coefficients of the median relation for solar activity dependence (\ref{eq:median_with_ssn}) with the ones from the exponential distance dependence (\ref{eq:exponential_function})
\begin{align}
	x_\text{med}(ssn,r) &= (a_\text{med} \cdot ssn + b_\text{med}) \cdot r^b
\end{align}

to get the combined model function $W''(x,ssn,r)$. And for the velocity $W_{II}''(x,ssn,r)$ with the double lognormal function.\\

%model errors
give model errors sizes/uncertainties (sigmas); $W''_\text{err}(x,ssn,r)$, $x_\text{med\_err}(ssn,r)$

@@

%SPP orbit
SPP orbit description, Venus flybys, within ecliptic, \citep{Fox2015}, (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_b_plot})\\

%SSN prediction
choosing SSN prediction source as input (SIDC + SWPC + previous cycle), (see Fig.~\ref{fig:SPP_orbit_predicted_SSN_overview_b_plot})\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_orbit_predicted_SSN_overview_b_plot.pdf}}
	\caption{SPP's solar distance during its mission time (top). Consecutive Venus flybys bring its perihelia nearer to the Sun. Actual and predicted SSN (bottom), i.e. SIDC 13-month smoothed monthly actual SSN, SIDC prediction, SWPC prediction and simply shifted SSN from previous cycle~24, together with two alternative trends of half and twice the amplitude.}
	\label{fig:SPP_orbit_predicted_SSN_overview_b_plot}
\end{figure}

Implementing the orbital distance data and predicted SSN for the mission time we get SPP's estimated solar wind environment.\\
A zoom into the first and the nearest perihelia shows... (Fig.~\ref{fig:SPP_perihelia_prediction_plot}).\\
\begin{figure}
	\resizebox{\hsize}{!}{\includegraphics{figures/SPP_perihelia_prediction_plot.pdf}}
	\caption{Estimated B-field median during the first and nearest perihelia in 2018 and 2024.}
	\label{fig:SPP_perihelia_prediction_plot}
\end{figure}

@@

velocity:\\
%critical surfaces
Alfvénic critical surface i.e. source surface (see Fox before 2.1)\\
in direction to the Sun is at $\sim$2.5~Rs the source surface (Schatten1969)\\
sonic and Alfvénic critical point positions (see \citet{Sittler1999})\\
sonic point and slow solar wind origin \citep{Sheeley1997}\\
approaching these regions, acceleration plays a role\\

@@

Comparison with existing near-Sun models (renew extrapolation figure to 1-column):\\

@@

%model limits
model limits (spherical coordinates):\\
- heliocentric distance range 0.29--0.98~au\\
- rotational symmetry\\
- confined to ecliptic +-7.2 degrees\\
model constrictions:\\
- solar distance dependency function\\
- frequency distribution functions\\
- neglected influence from heliolatitude variation\\

@@

all estimates outside these boundaries are extrapolations with large uncertainties.\\

@@

The use of hourly data instead of higher resolution data (less than 1~min) has no significant influence on the bulk results. The minutely OMNI data frequency distributions are almost congruent and differ only slightly at their extreme ends.\\
The many hourly Helios data points which contain only a few measurements, contribute with a larger scatter to the frequency distributions \citep[p.~491]{Schwenn1983}.\\

The solar wind parameters vary with solar distance as well as with latitudinal separation from the heliospheric current sheet (HCS).\\
The OMNI data is time-shifted to the nose of the Earth's bow shock. This leads to yearly solar distance variations of \SI{>2}{\percent} (cite?) as the Earth orbits the Sun. Furthermore, its orbit within the ecliptic leads to a yearly variation of \SI{+-7.2}{\degree} in heliospheric latitude.\\
The HCS's position in latitude is highly variable around the solar equator \citep[p.~127~ff.?]{Schwenn1990}.\\

Error estimation over the year (seasonal/monthly) -> we expect variations to be less than 5~\%\\

@@

list of results:\\
- empirical solar wind model for inner heliosphere within ecliptic\\
- low velocity at 0.0459~au\\
- slow/fast ratio SSN dependency\\
- application validity of lognormal distributions\\
--> B inversion of frequency distribution\\
---> magnetic field distribution's with distance increasing high value tail -> source are compression regions (why with density no increase?); look into Parker1958's B-field formula...\\
varying shape with distance is indicator for internal physical processes (mixing/turbulence...)\\

@@

@@

Further investigations should be done into structure extrapolations, outward extension of model to Mars...\\
